---
title: "Data Science With R-Project Healthcare - Simplilearn"
author: "Dheeraj Bharat Sethi"
date: "8/1/2020"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Project Statement

A nationwide survey of hospital costs conducted by the US Agency for 
Healthcare consists of hospital records of inpatient samples. 
The given data is restricted to the city of Wisconsin and relates to #patients in the age group 0-17 years. The agency wants to analyze the data to 
research on healthcare costs and their utilization.

First, lets import the required csv file, and get the feel of data, columns etc
We will be using various R basic functions to do that
```{r Hospital}
hospital_data<-read.csv(file.choose(),header = T)
names(hospital_data) #Get the columns
head(hospital_data) #First 6 rows
str(hospital_data) #Structure, datatype of each column
```

# Task 1
To record the patient statistics, the agency wants to find the 
age category of people who frequent the hospital and has the maximum expenditure.

So, here we need to find out the number of records grouped by AGE
As AGE is int variable, its better to convert that to a factor, and then use table, summary functions to get the count of rows based on AGE values

```{r Task 1}
#hospital_data$AGE<-as.factor(hospital_data$AGE)
levels(as.factor(hospital_data$AGE))
table(as.factor(hospital_data$AGE)) #here AGE 0 has max number, getting an idea
summary(hospital_data) #Summary of each variable in the dataset

#Now the max cost analysis, we need the aggregated values of TOTCHRG based on AGE, we can use aggregate function
cost_aggregate<-aggregate(TOTCHG~AGE,data = hospital_data,FUN = sum)
cost_aggregate
max(cost_aggregate) #678118
cost_aggregate[which.max(cost_aggregate$TOTCHG),] #Tells AGE=0 as maximum entry in aggregate
```

## Conclusion 1: 
Above output clearly shows the AGE=0 has 307 records, meaning, patients with AGE=0 are the most frequent visitors to the hospital_data.

## Conclusion 2: 
Aggregated expenditure of AGE=0 is the maximum, meaning, AGE=0 has the maximum costs.

Its time to PLOT these to visualize the observation
We can draw a Histogram, or a BarPlot
```{r Task 1 plot}
hist(hospital_data$AGE,breaks = nlevels(as.factor(hospital_data$AGE)),xlab = "Age",ylab = "Total Records",col = "green",freq = T,density = 100,border = 4)
barplot(table(hospital_data$LOS,hospital_data$AGE),xlab = "Age",ylab = "Total rows, Days Stayed stacked",col = "green",density = 100,border = 4)

```

# Task 2
In order of severity of the diagnosis and treatments and to find out the expensive treatments, the agency wants to find the diagnosis-related group that has maximum hospitalization and expenditure.

This is similar problem to Task 1, here we need to aggregate the Cost based on the diagnosis codes
```{r Task 2}
summary(as.factor(hospital_data$APRDRG)) #we can see code 640 has 267 records out of 500
cost_diag_aggregate<-aggregate(TOTCHG~APRDRG,data = hospital_data,FUN = sum)
cost_diag_aggregate
max(cost_diag_aggregate) #678118
cost_diag_aggregate[which.max(cost_diag_aggregate$TOTCHG),]
```

## Conclusion: 
Above output clearly shows the APRDG=640 has 267 records, meaning, and also the maximum costs value.

Its time to PLOT these to visualize the observation
We can draw a Histogram, or a BarPlot

``` {r Task 2 Plot}
hist(hospital_data$APRDRG,breaks = nlevels(as.factor(hospital_data$APRDRG)),xlab = "Diagnosis Code",ylab = "Sum of expenses",col = "green",freq = T,density = 100,border = 4)
#Histo chart shows the same results,
#Barplot won't be practical here, as it would take a lot of space to draw x bars. As can be seen with unique number of values of diagnosis codes
nlevels(as.factor(hospital_data$APRDRG))
#or
unique(hospital_data$APRDRG)
#We have 63 unique codes
```


# Task 3
To make sure that there is no malpractice, the agency needs to analyze if the race of the patient is related to the hospitalization costs.

Here, we have RACE - Categorical variable

Cost - Continuous variable

To find the relation between such combination, we can use ANOVA test

And we are doing the Hypothesis Testing 

Null Hypothesis H0: RACE has No Effect on Costs

Alternate Hypothesis Ha : RACE has Effect on Costs

To REJECT the H0, we need enough evidences, means, a striking difference between the both. 

In ANOVA, we have F-statistic

## Step 1

Data Wrangling/Cleaning

We need to clean up the entries with values NA in any column
## Step 2

Apply ANOVA test R function, on TOTCHG as Dependent, RACE as independent variable
```{r Task 3}
anyNA(hospital_data) #this tells TRUE
anyNA(hospital_data$RACE) #RACE has some NA entries
summary(hospital_data$RACE) # 1 NA value, we can omit this, as we have no way to predict this as we do for numerical variables
hospital_data<-na.omit(hospital_data)
anyNA(hospital_data) #Now we have no NA values
#ready to test ANOVA now
model_race_vost_aov<-aov(hospital_data$TOTCHG~hospital_data$RACE) 
summary(model_race_vost_aov)
```

## Conclusion
F-statistic is 0.164, and P-value (probability) is 0.686 > 0.05 (significance level), hence we CANNOT REJECT the NULL Hypothesis.

Means RACE HAS NO IMPACT ON COSTS


# Task 4
To properly utilize the costs, the agency has to analyze the severity of the hospital costs by age and gender for the proper allocation of resources.

Hospital wants to understand, whether The Cost Increases/decreases with AGE or GENDER, so we need to find the linear relation among these.

Using Linear Regression model, with TOTCHG as dependent (predicted) variable, GENDER, AGE as independent (predictor) variables
``` {r Task 4}
hospital_data$FEMALE<-as.factor(hospital_data$FEMALE)
summary(hospital_data$FEMALE) #almost similar number of MALES and FEMALES
fit_1<-lm(formula =  TOTCHG~AGE+FEMALE,data = hospital_data)
fit_1
summary(fit_1)
```
## Conclusion 
We see, p-values for AGE is much lower than significance level, but p-value for Gender is much closer to 0.05.

That means, AGE has bigger impact on costs, although GENDER also is significant for the model.

Coefficient of AGE is higher, means cost is positively increasing with age.

Coefficient of FEMALE 1 is negative, tells, Costs for female patients are lesser as compare to the Males

Lets now plot out model fit_1
```{r Task 4 Plot}
library(ggplot2)
ggplot(hospital_data,aes(y=TOTCHG,x=AGE,color=factor(FEMALE)))+geom_point()+stat_smooth(method="lm",se=FALSE)
#The plot clearly defines our regression model, female have lesser costs than males,
#and More the age, more the costs
```

# Task 5
Since the length of stay is the crucial factor for inpatients, the agency wants to find if the length of stay can be predicted from age, gender, and race.

This is similar problem as 3, here the predictor variables are AGE,GENDER,RACE Predicted variable is Length of stay.

Lets use linear regression to find the relation

```{r Task 5}
hospital_data$RACE<-as.factor(hospital_data$RACE)
hospital_data<-na.omit(hospital_data)
fit_los<-lm(data = hospital_data,formula = LOS~AGE+FEMALE+RACE)
fit_los
summary(fit_los)
```

## Conclusion
None of the predictor variables show significance to the model, 

All the p-values are high, thus, Accepting the Null Hypothesis

Concluding that Length of stay cannot be predicted by AGE, RACE, or GENDER


# Task 6
To perform a complete analysis, the agency wants to find the variable that mainly affects hospital costs.

Predict Costs, with all other independent variables.

```{r Task 6}
fit_costs_all<-lm(data = hospital_data,formula = TOTCHG~AGE+FEMALE+RACE+APRDRG+LOS)
fit_costs_all
summary(fit_costs_all)
```

## Conclusion
AGE, LOS, APRDRG affect the costs for hospitals, none other does.

LOS has positive relation with Costs.

Here, LOS is the continuous variable, we can compare this with costs as=>

With each increment in length of days stayed, the Cost/Charges increases by 742.97 units.


